An Effective Model of HL-60 Differentiation

### 2017-04-14 Paper_v2.tex
Edits to materials and methods

The signal initiation model was bistable with respect to ATRA induction (Fig. \ref{fig:model-fitting}C-D).
Phaseplane analysis predicted two stable steady-states when ATRA was present below a critical threshold (Fig. \ref{fig:model-fitting}C).
In the lower stable state, neither the signalsome nor cRaf-pS621 were present (thus, the differentiation program was deactivated).
However, at the high stable state, both the signalsome and cRaf-pS621 were present, allowing for sustained activation and differentiation.
Interestingly, when ATRA was above a critical threshold, only the activated state was accessible (Fig. \ref{fig:model-fitting}D).
To test these findings, we first identified the ATRA threshold. We exposed HL-60 cells to different ATRA concentrations for 72 hr (Fig. \ref{fig:model-fitting}E).
Morphological changes associated with differentiation were visible for ATRA $\geq$ 0.25 $\mu$M, suggesting the critical ATRA
threshold was near this concentration. Next, we conducted ATRA washout experiments to determine if activated cells remained activated in the absence of ATRA.
HL-60 cells locked into an activated state remained activated following ATRA withdraw (Fig. \ref{fig:model-predictions}).
This sustained activation resulted from reinforcing feedback between the signalsome and the MAPK pathway.
Thus, following activation, if we inhibited or removed elements from the signal initiation module we expected the siganlsome and MAPK signals to decay.
We simulated ATRA induced activation in the presence of kinase inhibitors, and without key circuit elements.
Consistent with experimental results using multiple MAPK inhibitors, ATRA activation in the presence of MAPK inhibitors lowered the steady-state value of signalsome (Fig. \ref{fig:model-predictions}A).
In the presence of BLR1, the signalsome and cRaf-pS621 signals were maintained following ATRA withdraw (Fig. \ref{fig:model-predictions}B, gray).
On the other hand, BLR1 deletion removed the ability of the circuit to maintain a sustained MAPK response following the withdraw of ATRA (Fig. \ref{fig:model-predictions}B, blue).
Lastly, washout experiments in which cells were exposed to 1$\mu$M ATRA for 24 hr, and then transferred to fresh media without ATRA,
confirmed the persistence of the self sustaining activated state for up to 144 hr (Fig. \ref{fig:model-predictions}C).
Thus, these experiments confirmed that reinforcing positive feedback likely drives the ATRA-induced differentiation program.
Next, we analyzed the ATRA-induced downstream gene expression program following signalsome and cRaf activation.

@@

\subsubsection*{Signaling model equations.}
The signal initiation, and integration modules required the level of cRaf-pS621 and ATRA-RXR/RAR (activated Trigger) as inputs.
However, our base model described only the abundance of inactive proteins e.g., cRaf or RXR/RAR but not the activated forms.
To address this issue, we estimated pseudo steady state approximations for the abundance of cRaf-pS621 and activated Trigger (shown generally as Eq \eqref{eqn-signaling-balances}).
The abundance of activated trigger ($x_{a,1}$) was estimated directly from the RXR/RAR abundance ($x_{u,1}$):
\begin{equation}
	x_{a,1} \sim x_{u,1}\left(\frac{\alpha\cdot{\mathrm{ATRA}}}{1+\alpha\cdot\mathrm{ATRA}}\right)
\end{equation}
where $\alpha$ denotes a gain parameter; $\alpha = 0.0$ if ATRA is less than a threshold, and $\alpha = 0.1$ if ATRA is greater than the differentiation threshold.
The abundance of cRaf-pS621 was estimated by making the pseudo steady state approximation on the cRaf-pS621 balance.
The abundance of an activated signaling species $i$ was given by:
\begin{equation}
	\frac{dx_{i}}{dt}  =  r_{+,i}\left(\mathbf{x},\mathbf{k}\right) - \left(\mu+k_{d,i}\right)x_{i}\qquad{i=1,\hdots,\mathcal{M}}\\
\end{equation}
The quantity $x_{i}$ denotes concentration of signaling species $i$, while $\mathcal{R}$ and $\mathcal{M}$ denote the number of signaling reactions and signaling species in the model,
respectively. The term $r_{+,i}\left(\mathbf{x},\mathbf{k}\right)$ denotes the rate of generation of activated species $i$,
while $\mu$ denotes the specific growth rate, and $k_{d,i}$ denotes the rate constant controlling the non-specific degradation of $x_{i}$.
We neglected deactivation reactions e.g., phosphatase activities.
We assumed that signaling processes were fast compared to gene expression; this allowed us to approximate the signaling balance as:
\begin{equation}
	x_{i}^{*} \simeq \frac{r_{+,i}\left(\mathbf{x},\mathbf{k}\right)}{\left(\mu+k_{d,i}\right)} \qquad{i=1,\hdots,\mathcal{M}}
\end{equation}
The generation rate was written as the product of a kinetic term ($\bar{r}_{+,i}$) and a control term ($v_{i}$).
The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
If rate $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010138}.
If a process had no modifying factors, $v_{j}=1$.
The kinetic rate of cRaf-pS621 generation $\bar{r}_{+,cRaf}$ was modeled as:
\begin{equation}
	\bar{r}_{+,cRaf} = k_{+,cRaf}x_{s}\left(\frac{x_{cRaf}}{K_{+,cRaf}+x_{cRaf}}\right)
\end{equation}where $x_{s}$ denotes the signalsome abundance, and $K_{+,cRaf}$ denotes a saturation constant governing cRaf-pS621 formation.
The formation of cRaf-pS621 was regulated by only a single factor, the abundance of MAPK inhibitor, thus $v_{+,cRaf}$ took the form:
\begin{equation}
	v_{+,cRaf} = \left(1-\frac{\mathrm{I}}{K_{D}+\mathrm{I}}\right)
\end{equation}where $\mathrm{I}$ denotes the abundance of the MAPK inhibitor, and $K_{D}$ denotes the inhibitor affinity.

\subsubsection*{Effective cell cycle arrest model.}
We formulated an effective model of the fraction of cells undergoing ATRA-induced cell cycle arrest, $\mathcal{A}$ as:
\begin{equation}\label{eq:a-eqn}
	\frac{d\mathcal{A}}{dt} = \epsilon_{1}\left(\mathrm{E2F}\right)^{\epsilon_{2}}
\end{equation}where $\epsilon_{i}$ were unknown parameters, and $\mathrm{E2F}$ denotes the abundance of the E2F transcription factor.
The parameters $\epsilon_{1}$ and $\epsilon_{2}$ were estimated from experimental cell cycle distribution measurements, by discretizing and \texttt{log} transforming Eq. \eqref{eq:a-eqn}:
\begin{eqnarray}
\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{1}\right) & = & \log{\dot{\mathcal{A}}_{21}} \\
\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{2}\right) & = & \log{\dot{\mathcal{A}}_{32}}
\end{eqnarray}
where subscripts denote the experimental time point (1 = 24 hr, 2 = 48 hr, and 3 = 72 hr), $\mathrm{E2F}_{i}$ denotes the $\mathrm{E2F}$ abundance at time point $i$,
and $\dot{\mathcal{A}}_{ij}$ denotes the time derivative of the fraction of arrested cells estimated between time point $i$ and $j$ (estimated from data).
We calculated candidate values for $\epsilon_{j}$ by solving the log-transformed system of equations for 10 biological replicates.

@@

\noindent\subsubsection*{Estimation of model parameters.}
We estimated parameters appearing in the mRNA and protein balances, and the abundance of polymerases and ribosomes, from estimates of transcription and translation rates,
the half-life of a typical mRNA and protein, and typical values for the copies per cell of RNA polymerase and ribosomes from literature (Table \ref{tbl:model-parameters}).
For the remaining parameters, e.g., the $W_{ij}$ appearing in the control laws,
or parameters appearing in the transfer functions $f_{dj}$, were estimated from the gene expression and signaling data sets discussed here.
The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
Lastly, we calculated the concentration for gene $G_{j}$ by assuming, on average, that a cell had two copies of each gene at any given time.
Thus, the bulk of our gene expression parameters were based directly upon literature values, and were not adjusted during model identification.
The values used for the characteristic transcription/translation parameters, degradation constants and
macromolecular copy number are given in the supplemental materials along with the specific formulas required to calculate all derived constants.

Signal and gene expression model parameters were estimated by minimizing the squared difference between simulations and experimental data set $j$:
\begin{equation}\label{eqn:objective-function}
	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2 + \left(\frac{{\mathcal{M}^{\prime}_{ij}}-\max{y_{ij}}}{{\mathcal{M}^{\prime}_{ij}}}\right)^{2}
\end{equation}
The terms $\hat{\mathcal{M}}_{ij}$ and $\hat{y}_{ij}$ denote scaled experimental observations and simulation outputs at time $i$ from training set $j$,
where $\mathcal{T}_{j}$ denoted the number of time points for data set $j$.
The first term in Eqn. \eqref{eqn:objective-function} quantified the relative simulation error.
We used immunoblot intensity measurements for model training.
Thus, we trained the model on the \emph{relative} change between bands within each data set.
Suppose we have the intensity of species $x$ at time $\{t_{1},t_{2},..,t_{n}\}$ in condition $j$.
The scaled value $0\leq\hat{\mathcal{M}}_{ij}\leq{1}$ is given by:
\begin{equation}\label{norm_exp_data}
\hat{\mathcal{M}}_{ij} = \left({\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}\right)/\left({\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}\right)
\end{equation}
where $\hat{\mathcal{M}}_{ij}=0$ and $\hat{\mathcal{M}}_{ij}=1$  describe the lowest (highest) intensity bands.
A similar scaling was used for the simulation output. The second term in the objective function ensured a realistic concentration scale was estimated by the model.
We set the highest intensity band to $\mathcal{M}^{\prime}_{ij} = 10$ [AU] for all simulations.
We minimized the total model residual $\sum_{j}E_{j}$ using heuristic optimization starting from a random initial parameter guess.

The signaling and gene expression model equations were implemented in the Julia programming language, and solved using the \texttt{ODE23s} routine of the ODE package \cite{Julia}.
The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
Model parameters estimated directly or derived from literature, or data presented in this study are given in Table \ref{tbl:model-parameters}.

@@

\section*{Author's contributions}
J.V and A.Y directed the study. R.T, H.J and J.C conducted the cell culture measurements.
J.V and W.D developed the reduced order HL-60 models and the parameter ensemble.
W.D analyzed the model ensemble, and generated figures for the manuscript.
The manuscript was prepared and edited for publication by W.D, A.Y and J.V.

>>>>

The signal initiation model was bistable with respect to ATRA induction (Fig. \ref{fig:model-fitting}C-D).
Phaseplane analysis predicted two stable steady-states when ATRA was present below a critical threshold, and only a single steady-state above the threshold (Fig. \ref{fig:model-fitting}C).
In the lower stable state, neither the signalsome nor cRaf-pS621 were present (thus, the differentiation program was deactivated).
However, at the high stable state, both the signalsome and cRaf-pS621 were present, allowing for sustained activation and differentiation.
Interestingly, when ATRA was above a critical threshold, only the activated state was accessible (Fig. \ref{fig:model-fitting}D).
To test these findings, we first identified the ATRA threshold. We exposed HL-60 cells to different ATRA concentrations for 72 hr (Fig. \ref{fig:model-fitting}E).
Morphological changes associated with differentiation were visible for ATRA $\geq$ 0.25 $\mu$M, suggesting the critical ATRA
threshold was near this concentration. Next, we conducted ATRA washout experiments to determine if activated cells remained activated in the absence of ATRA.
HL-60 cells locked into an activated state remained activated following ATRA withdraw (Fig. \ref{fig:model-predictions}).
This sustained activation resulted from reinforcing feedback between the signalsome and the MAPK pathway.
Thus, following activation, if we inhibited or removed elements from the signal initiation module we expected the siganlsome and MAPK signals to decay.
We simulated ATRA induced activation in the presence of kinase inhibitors, and without key circuit elements.
Consistent with experimental results using multiple MAPK inhibitors, ATRA activation in the presence of MAPK inhibitors lowered the steady-state value of signalsome (Fig. \ref{fig:model-predictions}A).
In the presence of BLR1, the signalsome and cRaf-pS621 signals were maintained following ATRA withdraw (Fig. \ref{fig:model-predictions}B, gray).
On the other hand, BLR1 deletion removed the ability of the circuit to maintain a sustained MAPK response following the withdraw of ATRA (Fig. \ref{fig:model-predictions}B, blue).
Lastly, washout experiments in which cells were exposed to 1$\mu$M ATRA for 24 hr, and then transferred to fresh media without ATRA,
confirmed the persistence of the self sustaining activated state for up to 144 hr (Fig. \ref{fig:model-predictions}C).
Thus, these experiments confirmed that reinforcing positive feedback likely drives the ATRA-induced differentiation program.
Next, we analyzed the ATRA-induced downstream gene expression program following signalsome and cRaf activation.

@@

\subsubsection*{Signaling model equations.}
The signal initiation, and integration modules required the abundance of cRaf-pS621 and ATRA-RXR/RAR (activated Trigger) as inputs.
However, our base model described only the abundance of inactive proteins e.g., cRaf or RXR/RAR but not the activated forms.
To address this issue, we estimated pseudo steady state approximations for the abundance of cRaf-pS621 and activated Trigger (shown generally as Eq \eqref{eqn-signaling-balances}).
The abundance of activated trigger ($x_{a,1}$) was estimated directly from the RXR/RAR abundance ($x_{u,1}$):
\begin{equation}
	x_{a,1} \sim x_{u,1}\left(\frac{\alpha\cdot{\mathrm{ATRA}}}{1+\alpha\cdot\mathrm{ATRA}}\right)
\end{equation}
where $\alpha$ denotes a gain parameter; $\alpha = 0.0$ if ATRA is less than a threshold, and $\alpha = 0.1$ if ATRA is greater than the differentiation threshold.
The abundance of cRaf-pS621 was estimated by making the pseudo steady state approximation on the cRaf-pS621 balance.
The abundance of an activated signaling species $i$ was given by:
\begin{equation}
	\frac{dx_{i}}{dt}  =  r_{+,i}\left(\mathbf{x},\mathbf{k}\right) - \left(\mu+k_{d,i}\right)x_{i}\qquad{i=1,\hdots,\mathcal{M}}\\
\end{equation}
The quantity $x_{i}$ denotes concentration of signaling species $i$, while $\mathcal{R}$ and $\mathcal{M}$ denote the number of signaling reactions and signaling species in the model,
respectively. The term $r_{+,i}\left(\mathbf{x},\mathbf{k}\right)$ denotes the rate of generation of activated species $i$,
while $\mu$ denotes the specific growth rate, and $k_{d,i}$ denotes the rate constant controlling the non-specific degradation of $x_{i}$.
We neglected deactivation reactions e.g., phosphatase activities.
We assumed that signaling processes were fast compared to gene expression; this allowed us to approximate the signaling balance as:
\begin{equation}
	x_{i}^{*} \simeq \frac{r_{+,i}\left(\mathbf{x},\mathbf{k}\right)}{\left(\mu+k_{d,i}\right)} \qquad{i=1,\hdots,\mathcal{M}}
\end{equation}
The generation rate was written as the product of a kinetic term ($\bar{r}_{+,i}$) and a control term ($v_{i}$).
The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
If rate $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010138}.
If a process had no modifying factors, $v_{j}=1$.
The kinetic rate of cRaf-pS621 generation $\bar{r}_{+,cRaf}$ was modeled as:
\begin{equation}
	\bar{r}_{+,cRaf} = k_{+,cRaf}x_{s}\left(\frac{x_{cRaf}}{K_{+,cRaf}+x_{cRaf}}\right)
\end{equation}where $x_{s}$ denotes the signalsome abundance, and $K_{+,cRaf}$ denotes a saturation constant governing cRaf-pS621 formation.
The formation of cRaf-pS621 was regulated by only a single factor, the abundance of MAPK inhibitor, thus $v_{+,cRaf}$ took the form:
\begin{equation}
	v_{+,cRaf} = \left(1-\frac{\mathrm{I}}{K_{D}+\mathrm{I}}\right)
\end{equation}where $\mathrm{I}$ denotes the abundance of the MAPK inhibitor, and $K_{D}$ denotes the inhibitor affinity.

\noindent\subsubsection*{Estimation of gene expression model parameters.}
We estimated parameters appearing in the mRNA and protein balances, the abundance of polymerases and ribosomes,
transcription and translation rates, the half-life of a typical mRNA and protein,
and typical values for the copies per cell of RNA polymerase and ribosomes from literature (Table \ref{tbl:model-parameters}).
The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
Lastly, we calculated the concentration for gene $G_{j}$ by assuming, on average, that a cell had two copies of each gene at any given time.
Thus, the bulk of our gene expression model parameters were based directly upon literature values, and were not adjusted during model identification.
However, the remaining parameters, e.g., the $W_{ij}$ appearing in the gene expression control laws,
or parameters appearing in the transfer functions $f_{dj}$, were estimated from the protein expression and signaling data sets discussed here.

Signaling and gene expression model parameters were estimated by minimizing the squared difference between simulations and experimental protein data set $j$.
We measured the squared difference in the scale, fold change and shape for protein $j$:
\begin{equation}\label{eqn:objective-function}
	E_{j}(\mathbf{k}) = \biggl(\mathcal{M}_{j}\left(t_{-}\right)-\hat{y}_{j}\left(t_{-},\mathbf{k}\right)\biggr)^2+\sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2 + \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\mathcal{M}^{\prime}_{ij}-y^{\prime}_{ij}(\mathbf{k})\biggr)^2
\end{equation}
The first term in Eqn. \eqref{eqn:objective-function} quantified the initial \textit{scale} error, directly before the addition of ATRA.
In this case, $\mathcal{M}_{j}\left(t_{-}\right)$ (the approximate concentration of protein $j$ before the addition of ATRA) was estimated from literature.
This term was required because the protein measurements were reported as the fold-change; thus, the data was normalized by a control
value measured before the addition of ATRA. However, the model operated on a physical scale.
The first term allowed the model to capture physically realistic changes following ATRA addition.
The second term quantified the difference in the \textit{fold-change} of protein $j$ as a function of time.
The terms $\hat{\mathcal{M}}_{ij}$ and $\hat{y}_{ij}$ denote the scaled experimental observations and simulation outputs (fold-change; protein normalized by control value directly before ATRA addition)
at time $i$ from protein $j$,
where $\mathcal{T}_{j}$ denoted the number of time points for data set $j$.
Lastly, the third term of the objective function measured the difference in the \textit{shape} of the measured and simulated protein levels.
The scaled value $0\leq\mathcal{M}^{\prime}_{ij}\leq{1}$ was given by:
\begin{equation}\label{norm_exp_data}
\hat{\mathcal{M}}_{ij} = \left({\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}\right)/\left({\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}\right)
\end{equation}
where $\mathcal{M}^{\prime}_{ij}=0$ and $\mathcal{M}^{\prime}_{ij}=1$  describe the lowest (highest) intensity bands.
A similar scaling was used for the simulation output.
We minimized the total model residual $\sum_{j}E_{j}$ using a heuristic direct-search optimization procedure, subject to box constraints on the parameter values,
starting from a random initial parameter guess. Each downhill step was archived and used for ensemble calculations.
The optimization procedure have been reported previously [REFHERE].

% Model parameters estimated directly or derived from literature, or data presented in this study are given in Table \ref{tbl:model-parameters
\subsubsection*{Estimation of an effective cell cycle arrest model.}
We formulated an effective N-order polynomial model of the fraction of cells undergoing ATRA-induced cell cycle arrest at time t, $\hat{\mathcal{A}}(t)$, as:
\begin{equation}\label{eq:a-eqn}
	\hat{\mathcal{A}}\left(t\right)\simeq a_{0}+\sum_{i=1}^{N-1}a_{i}\phi_{i}\left(\mathbf{p}(t),t\right)
\end{equation}where $a_{i}$ were unknown parameters, and $\phi_{i}\left(\mathbf{p}(t),t\right)$ denotes a basis function.
The basis functions were dependent upon the system state; in this study, we assumed $N = 4$ and basis functions of the form:
\begin{equation}
	\phi_{i}\left(\mathbf{p}(t),t\right) = \left(\frac{t}{T}+\frac{p21}{E2F}\Bigr|_{t}\right)^{(i-1)}
\end{equation}
The parameters $a_{0},\hdots,a_{3}$ were estimated directly from cell-cycle measurements (biological replicates) using least-squares.
% The abundances of the $p21$ and $E2F$ proteins were taken from the gene expression model solution. The least squares solution for the polynomial
% coefficients was given by:
% \begin{equation}
% 	\mathbf{a} = \left[\left(\mathbf{V}^{T}\mathbf{V}\right)^{-1}\mathbf{V}^{T}\right]\mathcal{A}(t)
% \end{equation}where $\mathbf{V}$ denotes the Vandermonde matrix constructed from the basis functions, and $\mathcal{A}(t)$ denotes the vector of
% measured G0 fractions following ATRA exposure.

\subsubsection*{Availability of model code.}
The signaling and gene expression model equations, and the parameter estimation procedure, were implemented in the Julia programming language.
The model equations were solved using the \texttt{ODE23s} routine of the ODE package \cite{Julia}.
The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.

@@

@@

\section*{Author's contributions}
J.V and A.Y directed the study. R.T, H.J, R.B and J.C conducted the cell culture measurements.
J.V, R.B, W.D, K.R and A.S developed the reduced order HL-60 models and the parameter ensemble.
W.D and J.V analyzed the model ensemble, and generated figures for the manuscript.
The manuscript was prepared and edited for publication by W.D, A.Y and J.V.
